---
title: "Loan Approval Data Project"
author: "Trillyon Earl, Owen Jefcoat"
date: "2025-03-01"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<!-- Libraries & Setup -->
```{r libs, results=FALSE, warning=FALSE, message=FALSE}
library(MASS)
library(car)
library(nnet)
library(dplyr)
library(leaps)
```

<!-- Data Clean-Up -->

We regressed on whether a loan was approved or denied, based on data about applicants:

Variables in the dataset:

- **status**: (factor) Whether the loan was approved (1 for approved, 0 for denied) 
- **age**: (numeric) Age of applicant
- **gender**: (factor) Gender of applicant (1 for female, 0 for male)
- **education**: (ordered factor) Maximum educational attainment (High School -> Associate -> Bachelor -> Master -> Doctorate) 
- **income**: (numeric) Annual income of applicant ($)
- **exp**: (numeric) Employment experience in years
- **HO**: (factor) Home ownership status ("RENT", "OWN", "MORTGAGE", "OTHER")
- **loan_amnt**: (numeric) Amount of the loan ($)
- **loan_intent**: (factor) Intent of the loan ("PERSONAL", "EDUCATION", "MEDICAL", "VENTURE", "HOMEIMPROVEMENT", "DEBTCONSOLIDATION")
- **loan_int_rate**: (numeric) Interest rate of the loan (%)
- **loan_percent_income**: (numeric) Loan amount as a percentage of annual income (%)
- **cred_hist_length**: (numeric) Amount of years of previous credit history
- **credit_score**: (numeric) Credit score of applicant 
- **prev_defaults**: (factor) Whether applicant had previously defaulted on loan (1 for yes, 0 for no)

```{r cleanup}
loan <- read.csv("./loan_data.csv")
loan <- as_tibble(loan)
loan <- loan |> mutate(
  person_gender = factor(person_gender, ordered = FALSE, levels = c("female", "male")),
  person_education = factor(person_education, ordered = TRUE, levels = c("High School", "Associate", "Bachelor", "Master", "Doctorate")),
  person_home_ownership = factor(person_home_ownership, ordered = FALSE, c("RENT", "OWN", "MORTGAGE", "OTHER")),
  loan_intent = factor(loan_intent, ordered = FALSE, c("PERSONAL", "MEDICAL", "EDUCATION", "VENTURE", "HOMEIMPROVEMENT", "DEBTCONSOLIDATION"))
)

#Variable naming and indicator creation
age <- loan$person_age; gender <- loan$person_gender; education <- loan$person_education; income <- loan$person_income; exp <- loan$person_emp_exp; HO <- loan$person_home_ownership; loan_amnt <- loan$loan_amnt; loan_intent <- loan$loan_intent; loan_int_rate <- loan$loan_int_rate; loan_percent_income <- loan$loan_percent_income; cred_hist_length <- loan$cb_person_cred_hist_length; credit_score <- loan$credit_score; prev_defaults <- loan$previous_loan_defaults_on_file; status <- loan$loan_status

#Gender indicators (male is baseline)
female <- ifelse(gender == "female", 1, 0)

#Education indicators (High School is Baseline)
associate <- ifelse(education == "Associate", 1, 0)
bachelor <- ifelse(education == "Bachelor", 1, 0)
master <- ifelse(education == "Master", 1, 0)
phd <- ifelse(education == "Doctorate", 1, 0)

#Home ownership indicators (OTHER is baseline)
rent <- ifelse(HO == "RENT", 1, 0)
mortgage <- ifelse(HO == "MORTGAGE", 1, 0)
own <- ifelse(HO == "OWN", 1, 0)

#Loan intent indicators (PERSONAL is baseline)
education <- ifelse(loan_intent == "EDUCATION", 1, 0)
medical <- ifelse(loan_intent == "MEDICAL", 1, 0)
venture <- ifelse(loan_intent == "VENTURE", 1, 0)
homeimprov <- ifelse(loan_intent == "HOMEIMPROVEMENT", 1, 0)
debtconsol <- ifelse(loan_intent == "DEBTCONSOLIDATION", 1, 0)

#Previous loan defaults on file indicator (No is baselines)
yesdefault <- ifelse(prev_defaults == "Yes", 1, 0)

loan_indic <- as_tibble(cbind(age, female, associate, bachelor, master, phd, income, exp, rent, mortgage, own, loan_amnt, education, medical, venture, homeimprov, debtconsol, loan_int_rate, loan_percent_income, cred_hist_length, credit_score, yesdefault, status))
```

We used the first 36000 observations as training data, and the last 9000 as testing data.

```{r splittrainingandtesting}
#Split training and testing data
train_indic <- loan_indic[1:36000,]
test_indic <- loan_indic[36001:45000,]
```

## Abstract
<!-- Explain The Data -->
 - Explain what the data columns are, what the different values of each column.
 - Explain what data we decided to exclude before any training.
 
```{r exclusions}
loan_logit <- glm(status ~ ., family=binomial(link = "logit"), data=train_indic)
#Leverage
Xhat <- as.matrix(train_indic[-14])
lev <- hat(Xhat)
levthreshold <- 2*(24/36000)
levindex <- which(lev > levthreshold)

#Cook's Distance
cd <- cooks.distance(loan_logit)
cdindex <- which(cd > 0.8)

#Studentized residuals
rstud <- rstudent(loan_logit)
rstudindex <- which(rstud > 2)

intersect(levindex, rstudindex)
intersect(rstudindex, cdindex)
intersect(levindex, cdindex)

#Removing influential point
train_indic_fixed <- train_indic[-32298,]
```
 
## Objective
<!-- State the objective of this project -->
  - Analyze loan approval data and determine which factors are more or less likely to lead to people being approved or denied for loans.
  
## Methods
<!-- What type of regession techniques did we use? -->
  
We used binary logistic regression:

```{r logit}
loan_logit_fixed <- glm(status ~ ., family=binomial(link = "logit"), data=train_indic_fixed)
summary(loan_logit_fixed)
```

<!-- What type of variable selection methods did we use? -->

We used VIF analysis, AIC (Backwards, Forwards, and Stepwise), and Mallows' Cp statistic

```{r varselection}
#VIF analysis w/ drop in deviance testing
vifs <- vif(loan_logit_fixed)
which(vifs >= 10)
loan_logit_minusAge <- glm(status ~ ., family=binomial(link = "logit"), data=train_indic_fixed[,-1])
loan_logit_minusExp <- glm(status ~ ., family=binomial(link = "logit"), data=train_indic_fixed[,-8])
loan_logit_minusHO <- glm(status ~ ., family=binomial(link = "logit"), data=train_indic_fixed[,c(-9,-10,-11)])
anova(loan_logit_minusAge, loan_logit_minusExp, loan_logit_minusHO, loan_logit_fixed, test="Chisq")

#Backwards AIC
#back_AIC <- stepAIC(loan_logit_fixed, direction = "backward", k=2)
Cpsel <- leaps(as.matrix(train_indic_fixed[,-23]), train_indic_fixed$status, method = "Cp", nbest=2)
Table <- cbind(Cpsel$which, Cpsel$size, Cpsel$Cp)
Table[which(Cpsel$Cp == min(Cpsel$Cp)),]
train_indic_selected <- train_indic_fixed[,c(-1,-2,-3,-4,-5,-6)]
loan_logit_selected <- glm(status ~ ., family=binomial(link = "logit"), data=train_indic_selected)
anova(loan_logit_selected, loan_logit_fixed, test="Cp")
```

So our model with minimal Cp includes income, HO, loan_amnt, loan_intent, loan_int_rate, loan_percent_income, cred_hist_length, credit_score, and yesdefault.

```{r predictions}
#Predictions
predictions <- predict(loan_logit_fixed, newdata=data.frame(test_indic[,-23]), type="response")
mean((predictions > 0.5 & test_indic[23] == 1) | (predictions <= 0.5 & test_indic[23] == 0))
Cppredictions <- predict(loan_logit_selected, newdata=data.frame(test_indic[,c(-1,-2,-3,-4,-5,-6)]), type = "response")
mean((Cppredictions > 0.5 & test_indic[23] == 1) | (Cppredictions <= 0.5 & test_indic[23] == 0))
```

<!-- What type of model selection method did we use? -->

## Analysis
  <!-- Analysis Introduction -->
  
## Conclusion
  -